---
title: "Cleaning Hylaeus Observation Data"
format: html
editor: visual
creator: Korey "Fox" Wetherell
---

# Introduction

This script will walk through cleaning raw observation data collected during Hylaeus (native bee) fieldwork. 

This process is designed for beginners, with clear steps and explanations for each package and function used. We are focusing on making qualitative observations more organized for mixed-methods research.


> 💡 **Tip:** While R can read `.xlsx` files with the `readxl` package, it's best practice to convert data to `.csv` for long-term reproducibility and stability. Excel files can become corrupted or change format, while `.csv` files are plain text and easier to version control or share.

# Loading Required Libraries

```{r}
# Install packages if needed (uncomment and run if necessary)
# install.packages("readxl")
# install.packages("dplyr")
# install.packages("stringr")
# install.packages("lubridate")

# Load the libraries


# Cleaned R script from Cleaning_Hylaeus_Data.qmd (Final Fixed Version with Type Handling)
# Use this for live demo or reproducible processing of Hylaeus observations

# Load necessary libraries
library(readxl)    # For reading Excel files
library(dplyr)     # For data manipulation
library(lubridate) # For working with dates and times
library(stringr)   # For keyword searches

# Although readxl works fine for Excel, converting to .csv is recommended for archival and reproducibility.

# Define the path to the Excel file (relative path from /Code/ folder)
excel_file <- "../Data/Raw_data/Kaʻena_Hylaeus Data Observations (Grouped).xlsx"

# Step 1: Get sheet names
sheet_names <- excel_sheets(excel_file)

# Step 2: Convert sheet names (assumed to be in MMDDYY format) to proper Date
parsed_dates <- suppressWarnings(mdy(sheet_names))
bad_sheets <- is.na(parsed_dates)
print(sheet_names[bad_sheets])  # Optional: inspect bad sheet names
sheet_names <- sheet_names[!bad_sheets]
sheet_dates <- parsed_dates[!bad_sheets]

# Step 3: Read in all sheets, tagging each observation with its respective date
data_list <- lapply(seq_along(sheet_names), function(i) {
  read_excel(excel_file, sheet = sheet_names[i]) %>%
    mutate(
      Date = sheet_dates[i],
      Timestamp = as.character(Timestamp),
      `Temperature (°F)` = as.numeric(`Temperature (°F)`),
      `Wind Direction` = as.character(`Wind Direction`),
      `Wind Notes` = as.character(`Wind Notes`)
    )
})
    mutate(
      Date = sheet_dates[i],
      Timestamp = as.character(Timestamp)  # normalize column type
    )
})

# Step 4: Combine into a single dataframe
observations_raw <- bind_rows(data_list)

# Step 5: Combine Date and Timestamp columns into one unified Datetime column
observations_clean <- observations_raw %>%
  mutate(
    Datetime = parse_date_time(paste(Date, Timestamp), orders = "ymd IMp")
  )

# Step 6: Create logical columns for keyword searches
observations_clean <- observations_clean %>%
  mutate(
    Photo_Taken = if_else(str_detect(tolower(paste(Detail, Additional_Notes)), "photo|filmed|picture|video"), TRUE, FALSE),
    Species_Sighted = if_else(str_detect(tolower(paste(Detail, Additional_Notes)), "sighting|seen|observed|found"), TRUE, FALSE),
    Observation_Noted = if_else(str_detect(tolower(paste(Detail, Additional_Notes)), "observation|observed|noted|detected"), TRUE, FALSE)
  )

# Step 7: Reorder columns for clarity
observations_clean <- observations_clean %>%
  select(Datetime, Date, Timestamp, Category, Detail, Additional_Notes,
         Photo_Taken, Species_Sighted, Observation_Noted,
         `Temperature (°F)`, `Wind Direction`, `Wind Notes`)

# Step 8: Create directory for output if it doesn’t exist
dir.create("../Data/Processed_data", recursive = TRUE, showWarnings = FALSE)

# Step 9: Save outputs
write.csv(observations_clean, "../Data/Processed_data/Hylaeus_Observations_Cleaned.csv", row.names = FALSE)
saveRDS(observations_clean, "../Data/Processed_data/Hylaeus_Observations_Cleaned.rds")

library(readxl)    # For reading Excel files
library(dplyr)     # For data manipulation
library(stringr)   # For finding keywords in text
library(lubridate) # For working with dates and times
```

# Setting the Working Directory

This project uses **relative paths**, which means R looks for files starting from the current working directory.

Before running the code, make sure your working directory is set to the root folder of the project.  
You can check your current working directory by running:

```{r}
getwd()
```

If needed, you can list files in your working directory using:

```{r}
list.files()
```

This is helpful to double-check that your folders and files are organized correctly.

The Excel file should be stored at:

```
Data/Raw_data/Kaʻena_Hylaeus Data Observations (Grouped).xlsx
```

If your working directory is correct, the loading step below will work without any changes.

# Load the Raw Data

```{r}
# Load all sheets from the Excel file
# Cleaned R script from Cleaning_Hylaeus_Data.qmd (Final Fixed Version with Type Handling)
# Use this for live demo or reproducible processing of Hylaeus observations

# Load necessary libraries
library(readxl)    # For reading Excel files
library(dplyr)     # For data manipulation
library(lubridate) # For working with dates and times
library(stringr)   # For keyword searches

# Although readxl works fine for Excel, converting to .csv is recommended for archival and reproducibility.

# Define the path to the Excel file (relative path from /Code/ folder)
excel_file <- "../Data/Raw_data/Kaʻena_Hylaeus Data Observations (Grouped).xlsx"

# Step 1: Get sheet names
sheet_names <- excel_sheets(excel_file)

# Step 2: Convert sheet names (assumed to be in MMDDYY format) to proper Date
parsed_dates <- suppressWarnings(mdy(sheet_names))
bad_sheets <- is.na(parsed_dates)
print(sheet_names[bad_sheets])  # Optional: inspect bad sheet names
sheet_names <- sheet_names[!bad_sheets]
sheet_dates <- parsed_dates[!bad_sheets]

# Step 3: Read in all sheets, tagging each observation with its respective date
data_list <- lapply(seq_along(sheet_names), function(i) {
  read_excel(excel_file, sheet = sheet_names[i]) %>%
    mutate(
      Date = sheet_dates[i],
      Timestamp = as.character(Timestamp),
      `Temperature (°F)` = as.numeric(`Temperature (°F)`),
      `Wind Direction` = as.character(`Wind Direction`),
      `Wind Notes` = as.character(`Wind Notes`)
    )
})
    mutate(
      Date = sheet_dates[i],
      Timestamp = as.character(Timestamp)  # normalize column type
    )
})

# Step 4: Combine into a single dataframe
observations_raw <- bind_rows(data_list)

# Step 5: Combine Date and Timestamp columns into one unified Datetime column
observations_clean <- observations_raw %>%
  mutate(
    Datetime = parse_date_time(paste(Date, Timestamp), orders = "ymd IMp")
  )

# Step 6: Create logical columns for keyword searches
observations_clean <- observations_clean %>%
  mutate(
    Photo_Taken = if_else(str_detect(tolower(paste(Detail, Additional_Notes)), "photo|filmed|picture|video"), TRUE, FALSE),
    Species_Sighted = if_else(str_detect(tolower(paste(Detail, Additional_Notes)), "sighting|seen|observed|found"), TRUE, FALSE),
    Observation_Noted = if_else(str_detect(tolower(paste(Detail, Additional_Notes)), "observation|observed|noted|detected"), TRUE, FALSE)
  )

# Step 7: Reorder columns for clarity
observations_clean <- observations_clean %>%
  select(Datetime, Date, Timestamp, Category, Detail, Additional_Notes,
         Photo_Taken, Species_Sighted, Observation_Noted,
         `Temperature (°F)`, `Wind Direction`, `Wind Notes`)

# Step 8: Create directory for output if it doesn’t exist
dir.create("../Data/Processed_data", recursive = TRUE, showWarnings = FALSE)

# Step 9: Save outputs
write.csv(observations_clean, "../Data/Processed_data/Hylaeus_Observations_Cleaned.csv", row.names = FALSE)
saveRDS(observations_clean, "../Data/Processed_data/Hylaeus_Observations_Cleaned.rds")

# Check the first few rows to confirm it loaded properly
head(observations_raw)
```

# Initial Inspection

```{r}
# Peek at the data structure
str(observations_raw)

# Check the column names
colnames(observations_raw)
```

# Cleaning Steps

## 1. Create Datetime Column

```{r}
# Combine Date and Timestamp into one Datetime field
observations_clean <- observations_raw %>%
  mutate(
    Datetime = parse_datetime(paste(Date, Timestamp), format = "%m%d%y %I:%M %p")
  )

# Check if the Datetime column looks correct
head(observations_clean$Datetime)
```

**Explanation:** Combining date and time into one column makes it easier to sort and filter observations later.

## 2. Create Boolean Flags from Text

```{r}
# Create logical columns based on keyword searches
observations_clean <- observations_clean %>%
  mutate(
    Photo_Taken = if_else(str_detect(tolower(paste(Detail, Additional_Notes)), "photo|filmed|picture|video"), TRUE, FALSE),
    Species_Sighted = if_else(str_detect(tolower(paste(Detail, Additional_Notes)), "sighting|seen|observed|found"), TRUE, FALSE),
    Observation_Noted = if_else(str_detect(tolower(paste(Detail, Additional_Notes)), "observation|observed|noted|detected"), TRUE, FALSE)
  )

# Confirm new columns are added
head(observations_clean[, c("Photo_Taken", "Species_Sighted", "Observation_Noted")])
```

**Explanation:** Here, we combine `Detail` and `Additional_Notes` and search for keywords that suggest a photo, sighting, or general observation.

- `tolower()` makes the text lowercase to avoid missing matches like "Photo" vs "photo".
- `str_detect()` checks if any keywords are present.

## 3. Handle Missing Values

```{r}
# Summarize temperature to see missing values
summary(observations_clean$`Temperature (°F)`)

# Check missing Wind Notes and Wind Direction
sum(is.na(observations_clean$`Wind Notes`))
sum(is.na(observations_clean$`Wind Direction`))
```

**Explanation:** We are not filling or deleting missing values here — documenting them preserves important information about observation gaps.

## 4. Reorder Columns for Clarity

```{r}
# Reorganizing the columns to prioritize key fields
observations_clean <- observations_clean %>%
  select(Datetime, Date, Timestamp, Category, Detail, Additional_Notes,
         Photo_Taken, Species_Sighted, Observation_Noted,
         `Temperature (°F)`, `Wind Direction`, `Wind Notes`)

# Preview the reorganized data
head(observations_clean)
```

# Saving Cleaned Data

```{r}
# Save as CSV
write.csv(observations_clean, "Data/Processed_data/Hylaeus_Observations_Cleaned.csv", row.names = FALSE)

# Save as RDS (R's internal data format)
saveRDS(observations_clean, "Data/Processed_data/Hylaeus_Observations_Cleaned.rds")
```

**Explanation:**
- `.csv` is easy to open and share.
- `.rds` preserves R-specific formatting and is better for analysis inside R later.

# Data Dictionary

| Variable | Description |
|:--|:--|
| Datetime | Combined date and time of observation |
| Date | Date of observation (from sheet name) |
| Timestamp | Time of observation (as recorded) |
| Category | General category of observation (e.g., Fauna, Environmental Conditions) |
| Detail | Specific description of the observation |
| Additional_Notes | Extra notes recorded in the field |
| Photo_Taken | TRUE if photo or filming mentioned |
| Species_Sighted | TRUE if a species was sighted or mentioned |
| Observation_Noted | TRUE if an observation was specifically noted |
| Temperature (°F) | Recorded temperature (if taken) |
| Wind Direction | Recorded wind direction (if taken) |
| Wind Notes | Extra notes about wind patterns |


---

# Final Notes

- Always **clean data programmatically**, not manually, to preserve reproducibility.
- Missing values are natural in field observations; documenting them is important.
- Logical (TRUE/FALSE) flags make it easier to group and summarize types of observations later.
- **Good habits:** Check your work after big steps by using `head()`, `str()`, and `list.files()` to confirm results.

Good luck with your Hylaeus research! 

